An observational study of papers published in _PLOS ONE_ and studies posted to a trial registry.

### 2021-05-24 manuscript/asa_template/asa_template.Rmd

The statistical methods section for the ANZCTR data was missing for some studies and we examined if there were particular studies where this section was more likely to be missing. 
We used four independent variables of date, study type (observational or interventional), number of funders and target sample size. 
We used a logistic regression model fitted using a Bayesian paradigm. A small number of sections were labelled as "Not applicable", "Nil" or "None" and we changed these to missing.

@@

To address these limitations, NMF deals with high-dimensional data by mapping it to a lower-dimensional space. This mapping this achieved by approximating $P$ with two factor matrices: \( W \in R^{M \times g} \) and \( H \in R ^{N \times g} \) [@aggarwal2012mining]. The number of subgroups of common topic inferred from the data is represented by $g$, leading to the approximation:
\begin{equation}
\label{eq:approxPNMF}
 P \approx WH^{T}.
\end{equation}

@@

Following estimation, \textit{H} contains the information regarding topic membership for all studies. In our case, topic membership $(1,\ldots,g)$ for a statistical methods section is inferred from the maximum coefficient value in the corresponding row of $H$.

For our two datasets, we applied NMF varying $g$ from 1 to 50. Clustering solutions were assessed using silhouette score, within-cluster dispersion and between-cluster dispersion.

\subsection{Within topic analysis}

<!-- Clustering results were visualised by word clouds that summarised frequently occurring terms associated with membership to different topics.  -->

Within the _PLOS ONE_ dataset, statistical methods sections with high topic coherence scores were used to investigate evidence of boilerplate text, compared with other studies assigned to the same topic. Comparisons were carried out by calculating Jaccard similarity scores between studies at the sentence level. We chose the Jaccard score as an intuitive measure, which summarises the similarity between two sentences by the number of words common to each sentence (intersection), divided by the number of words that appeared in either the strongest match or the sentence being compared. Boilerplate text was defined based on word count (plus or minus 3 words) and a Jaccard score of 0.9 or higher.

Results were transformed to lower case for the clustering, but examples are given using the original capitalisation.

<!-- \subsection{Missing statistical methods sections} -->

@@

## Included studies

### PLOS ONE

@@

@@

Searching with the _PLOS ONE_ API returned `r format(n.api,big.mark=',')` results (DOIs) based on chosen search terms (\autoref{fig:plos-flowchart}). After partial matching, `r format(n.include,big.mark=',')` (`r perc.include`%) statistical methods sections were identified. In the final sample, 95,518 (85%) DOIs returned an exact match against common section headings: 64,133 for 'statistical analysis', 13,380 for 'statistical analyses' and 13,627 for 'data analysis'. For our random sample of DOIs that did not meet the partial matching criteria, initial search terms appeared in Introduction (number), Results (number) only.  xxx appeared in the Methods section without a section heading and xxx had a non-standard section heading. [TODO].

Search results varied by journal volume (Figure \ref{fig:plos-n}A). The total number of API search results peaked at volumes 8 (n = `r format(total.records[[2]] %>% filter(volume==8) %>% pull(API),big.mark=',')`) and 9 (n = `r format(total.records[[2]] %>% filter(volume==8) %>% pull(API),big.mark=',')`) in years 2013 and 2014. This trend aligned with the total number of papers published in _PLOS ONE_ over the same period. The percentage of records that included a statistical methods section by volume based on our proposed matching criteria varied between `r round(min.perc.stats[['percentage']])`% (volume `r min.perc.stats[['volume']]`) and `r round(max.perc.stats[['percentage']])`% (volume `r max.perc.stats[['volume']]`).

All DOIs included Biology and life sciences (n = `r format(n.biology,big.mark=',')`), Earth sciences
(n = `r format(n.earth,big.mark=',')`) and/or Computer and information sciences (n = `r format(n.computer,big.mark=',')`) in their top 3 subject classifications (Figure \ref{fig:plos-n}B).

@@

<!-- Increasing the number of clusters decreased cluster quality based on global goodness-of-fit measures (Supplementary Figure 1), with average silhouette score and within-cluster dispersion leveling off around 20 clusters. This indicated that the data comprised one large, heterogeneous cluster and multiple smaller clusters. -->

@@

The topic clouds based on ten clusters are in Figure \ref{fig:plos-clouds}. Frequently occurring words reflected the use of statistical software (topics 3 and 5), descriptive statistics (topic 6), group based hypothesis testing (topics 1 and 4) and definitions of statistical significance (topics 1 and 9). Statistical methods sections related to regression (topic 2) and meta-analysis (topic 7) were also identified.

@@

Topics related to statistical software differentiated between Prism GraphPad (topic 3: n = `r filter(ftab.plos,topic_id=='Topic 3') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 3') %>% pull(perc)`%) and SPSS (topic 5: n = `r filter(ftab.plos,topic_id=='Topic 5') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 5') %>% pull(perc)`%) . Manual review of results showed that nine out of the ten top matching DOIs for topic 3 stated the use of Prism GraphPad, but did not specify which statistical methods were used. Top matching sections for topic 5 included information on SPSS version numbers and definitions of statistical significance.

@@

```{r}
jaccard.matches = read.xlsx('../../results/plos.jaccard_matches.xlsx')
ftab = jaccard.matches %>% mutate(Topic = paste('Topic',topic)) %>%
  select(Topic,sentences,matches) %>%  arrange(Topic,-matches) %>%
  rename('Statistical methods text' = sentences,
         'Number of matching studies' = matches)
knitr::kable(ftab,format = "latex", caption='Example boilerplate text from PLOS ONE dataset')
```

Definitions of statistical significance featured strongly in topic 1 (n = `r filter(ftab.plos,topic_id=='Topic 1') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 1') %>% pull(perc)`%) and topic 9 (n = `r filter(ftab.plos,topic_id=='Topic 9') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 9') %>% pull(perc)`%), combined with text describing hypothesis testing for comparing differences between two groups. Topic 1 reflected applications of Student's t-test assuming a 5% level of statistical significance. Topic 9 referenced similar methods combined with multiple thresholds for declaring statistical significance by asterisk: ``$^{*}p<0.05$, $^{**}p<0.01$ and $^{***}p<0.001$'', a practice that has been criticised [@Wasserstein2019].

Group-based hypothesis testing was a recurring theme across topics, with text descriptions varying based on method(s) used. One-way analysis of variance also featured strongly in topic 4 (n = `r filter(ftab.plos,topic_id=='Topic 4') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 4') %>% pull(perc)`%), combined with common methods for performing post-hoc multiple comparisons.  Frequently occurring words in topic 6 (n = `r filter(ftab.plos,topic_id=='Topic 6') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 6') %>% pull(perc)`%) reflected mentions of descriptive statistics for summarising continuous variables, for example: ``Data are expressed as means ± standard error of the mean (SEM).'' Sections assigned to this topic appeared to be expanded versions of topics 1 and 9. Examples of boilerplate text were in the form of descriptive statistics followed by hypothesis testing, for example, Student's t-test, Mann Whitney U or one-way analysis of variance.

@@

Examples of boilerplate text for selected topics are presented in Table \ref{tab:plos-example-boilerplate}.

@@

We downloaded 28,008 studies. The numbers of excluded studies are shown in Figure \ref{fig:anzctr-missing}. Of the 12,700 included studies, 9,523 (75%) had a statistical methods section. The median length of sections was 129 words with an inter-quartile range of 71 to 219 words.

We examined if four study characteristics were associated with a missing statistics section. The odds ratios and 95% credible intervals are in Table \ref{tab:anzctr-missing-odds}. Observational studies were less likely to have a missing methods section compared with interventional studies. Missing sections became less likely over time. Studies with more funders and a larger target sample size were less likely to have a missing methods section.

\begin{table}[]
\centering
\caption{Regression results for factors associated with missing statistical methods sections in ANZCTR}
\label{tab:anzctr-missing-odds}
\begin{tabular}{lcc}
\hline
Variable & Odds ratio & 95$\%$ CI \\
\hline
Study type $=$ Observational & 0.78 & (0.69, 0.89) \\
Date (per year) & 0.90 & (0.88, 0.91) \\
Number of funders & 0.80 & (0.74, 0.86) \\
Target sample size (per doubling) & 0.90 & (0.88, 0.92)\\
\hline
\end{tabular}
\end{table}

```{r, echo=FALSE, fig.align='center', out.width="70%", fig.cap="\\label{fig:anzctr-missing}ANZCTR search results."}

knitr::include_graphics("figures/excluded_anzctr_missing.jpg")

@@

Finally, targeted searches for the term 'statistician' revealed instances across all topics where no methods were specified. Instead, all analyses were the responsibility of an appointed statistician, for example, "A statistician employed by hospital was used" and ``Pilot study at this point will use a statistician professionally to determine sample size calculations as required''.

>>>>

Statistical methods sections were missing for some studies downloaded from ANZCTR, including a small number of sections labelled as "Not applicable", "Nil" or "None". We therefore examined if there were particular studies where this section was more likely to be missing. Analysis considered a logistic regression model estimated in the Bayesian framework [@INLA; www.r-inla.org], with missing statistical methods section (yes/no) as the dependent variable. Date, study type (observational or interventional), number of funders and target sample size were used as independent variables. Results were reported as odds ratios with 95% credible intervals (CI).

@@

To address these limitations, NMF deals with high-dimensional data by mapping it to a lower-dimensional space. This mapping this achieved by approximating $P$ with two factor matrices: \( W \in R^{M \times g} \) and \( H \in R ^{N \times g} \) [@aggarwal2012mining], such that $P \approx WH^{T}$. The number of subgroups of common topic inferred from the data is given by $g$.

@@

Following estimation, \textit{H} contains the information regarding topic membership for all studies. In our case, topic membership $(1,\ldots,g)$ for a statistical methods section is inferred from the maximum coefficient value in the corresponding row of $H$, also known as the topic coherence score. For our two datasets, we applied NMF with $g=10$.

\subsection{Content analysis}

Clustering results were visualised by word clouds to summarise frequently occurring terms associated with topic membership. Whilst these results were useful for inferring common themes across statistical methods sections, they did not indicate evidence of boilerplate text beyond the use of common terms. Follow up analysis therefore considered similarities in text between sections assigned to the same topic.

We used the Jaccard similarity index to compare the statistical methods section with the highest topic coherence score with all other sections assigned to the same topic. Indices were calculated at the sentence level, to account for instances of boilerplate text within larger methods sections. We chose the Jaccard index as an easy to interpret measure, which summarises the similarity between two sentences by the number of words common to each sentence (intersection), divided by the number of words that appeared in either sentence (union). Instances of boilerplate text were defined by a Jaccard index of 0.9 or higher and a difference in word count of plus or minus three words.

@@

## PLOS ONE

@@

```{r, echo=FALSE, fig.show='hold',fig.align='center', out.width="49%", fig.cap="\\label{fig:consort-diagrams}Search results for PLOS ONE (left) and ANZCTR (right)."}

knitr::include_graphics(c("figures/excluded_plosone.jpg","figures/excluded_anzctr_missing.jpg"))

@@

API searches returned `r format(n.api,big.mark=',')` results (DOIs) (\autoref{fig:consort-diagrams}). After partial matching, `r format(n.include,big.mark=',')` (`r perc.include`%) statistical methods sections were identified. In the final sample, 95,518 (85%) DOIs returned an exact match against common section headings: 64,133 for 'statistical analysis', 13,380 for 'statistical analyses' and 13,627 for 'data analysis'.

Search results varied by journal volume (Figure \ref{fig:plos-n}A). The total number of API search results peaked at volumes 8 (n = `r format(total.records[[2]] %>% filter(volume==8) %>% pull(API),big.mark=',')`) and 9 (n = `r format(total.records[[2]] %>% filter(volume==8) %>% pull(API),big.mark=',')`) in years 2013 and 2014. This trend aligned with the total number of papers published in _PLOS ONE_ over the same period. The percentage of records that included a statistical methods section by volume based on our proposed matching criteria varied between `r round(min.perc.stats[['percentage']])`% (volume `r min.perc.stats[['volume']]`) and `r round(max.perc.stats[['percentage']])`% (volume `r max.perc.stats[['volume']]`). All DOIs included Biology and life sciences (n = `r format(n.biology,big.mark=',')`), Earth sciences
(n = `r format(n.earth,big.mark=',')`) and/or Computer and information sciences (n = `r format(n.computer,big.mark=',')`) in their top 3 subject classifications (Figure \ref{fig:plos-n}B).

@@

For studies excluded based on section headings, initial search terms appeared in Introduction (number), Results (number) only.  xxx appeared in the Methods section without a section heading and xxx had a non-standard section heading. [TODO].

@@

The topic clouds for ten clusters are in Figure \ref{fig:plos-clouds}. Frequently occurring words reflected the use of statistical software (topics 3 and 5), descriptive statistics (topic 6), group based hypothesis testing (topics 1 and 4) and definitions of statistical significance (topics 1 and 9). Statistical methods sections related to regression (topic 2) and meta-analysis (topic 7) were also identified. Examples of boilerplate text for selected topics are presented in Table \ref{tab:plos-example-boilerplate}.

@@

Topics related to statistical software differentiated between Prism GraphPad (topic 3: n = `r filter(ftab.plos,topic_id=='Topic 3') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 3') %>% pull(perc)`%) and SPSS (topic 5: n = `r filter(ftab.plos,topic_id=='Topic 5') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 5') %>% pull(perc)`%). Within topic 3, the sentence "GraphPad Prism (Graphpad Software, San Diego, CA) was used for all analyses" was found in 1,941 studies. Similar phrasing was identified for topic 5: "SPSS software version 17.0 (SPSS, Chicago, IL, USA) was used for statistical analysis", with similar text found in 441 studies.

@@

```{r}
jaccard.matches = read.xlsx('../../results/plos.jaccard_matches.xlsx')
ftab = jaccard.matches %>% mutate(Topic = paste('Topic',topic)) %>%
  select(Topic,sentences,matches) %>%  arrange(Topic) %>%
  rename('Statistical methods text' = sentences,
         'Matching DOIs' = matches)
knitr::kable(ftab,format = "latex", caption='Example boilerplate text from PLOS ONE dataset')
```

Definitions of statistical significance featured strongly in topic 1 (n = `r filter(ftab.plos,topic_id=='Topic 1') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 1') %>% pull(perc)`%) and topic 9 (n = `r filter(ftab.plos,topic_id=='Topic 9') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 9') %>% pull(perc)`%). Topic 1 reflected applications of two-tailed and unpaired Student's t-tests, however, instances of boilerplate text emphasised the 5% level of statistical significance (\ref{tab:plos-example-boilerplate}). In contrast, Topic 9 focused on multiple thresholds for declaring statistical significance by asterisk: ``$^{*}p<0.05$, $^{**}p<0.01$ and $^{***}p<0.001$'', a practice that has been criticised [@Wasserstein2019].

Group-based hypothesis testing was a recurring theme across topics, with text descriptions varying based on method(s) used. One-way analysis of variance also featured strongly in topic 4 (n = `r filter(ftab.plos,topic_id=='Topic 4') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 4') %>% pull(perc)`%), combined with common methods for performing post-hoc multiple comparisons. Based on the Jaccard index, 1 in 5 studies were matched to the sentence ``Significant differences were determined using analysis of variance (ANOVA) follwo by Tukey post-hoc tests for multiple comparisons''. Frequently occurring words in topic 6 (n = `r filter(ftab.plos,topic_id=='Topic 6') %>% pull(DOIs)`; `r filter(ftab.plos,topic_id=='Topic 6') %>% pull(perc)`%) reflected mentions of descriptive statistics for summarising continuous variables, namely means with standard deviations or standard errors.

@@

@@

We downloaded 28,008 studies. The numbers of excluded studies are shown in Figure \ref{fig:consort-diagrams}. Of the 12,700 included studies, 9,523 (75%) had a statistical methods section. The median length of sections was 129 words with an inter-quartile range of 71 to 219 words.

Odds ratios and 95% credible intervals for study characteristics included in the logistic regression model are presented in Table \ref{tab:anzctr-missing-odds}. Observational studies were less likely to have a missing statistical methods section compared with interventional studies. Missing sections became less likely over time. Studies with more funders and a larger target sample size were less likely to have a missing statistical methods section.

\begin{table}[]
\centering
\caption{Logistic regression results for study characteristics associated with missing statistical methods sections in ANZCTR}
\label{tab:anzctr-missing-odds}
\begin{tabular}{lcc}
\hline
Variable & Odds ratio & 95$\%$ CI \\
\hline
Study type $=$ Observational & 0.78 & (0.69, 0.89) \\
Date (per year) & 0.90 & (0.88, 0.91) \\
Number of funders & 0.80 & (0.74, 0.86) \\
Target sample size (per doubling) & 0.90 & (0.88, 0.92)\\
\hline
\end{tabular}
\end{table}

@@

Since studies included in this dataset outlined planned analyses, we suspected that some studies did not specify statistical methods because they had yet to consult with a statistician. Targeted searches for the term "statistician" across all topics returned xx matching records, with examples including "A statistician employed by hospital was used" and ``Pilot study at this point will use a statistician professionally to determine sample size calculations as required''.
